### 2016-11-08 datamining/summery.tex
added more graph stuff

\begin{center}
     \Large{\underline{Datamining}} \\
\end{center}
\section{Classification trees}
\subsection{Building classification trees}
The objective of tree construction is to finally obtain nodes that are pure,
in the sense that they contain cases of a single class only.
\subsubsection{Impurity and quality of a split}
Impurity is a function:
\[i(t)=\phi(p_1,p_2,...,p_J)\]
where $p_j(j=1,...,J)$ is the relative frequency of the $J$ different classes.
$\phi$ is the impurity function with the following constraints:
\begin{enumerate}
  \item An impurity measure of a node should be at maximum when the observations
    are distributed evenly over all classes in that node.
  \item It should be at a minimum when all observations belong to a single class.
  \item $\phi$ is a symmetric function of $p_1,...,p_J$.
    a symmetric function of n variables is one whose value at any n-tuple of
    arguments is the same as its value at any permutation of that n-tuple.
\end{enumerate}
The quality of a split is defined as follows:
\[\Delta i(s,t)=i(t)-\phi(l)i(l)-\phi(r)i(r)\]
Where $\phi$ is a function that determines the proportion of cases for a node
and $l$ and $r$ are the left and right nodes of $t$.
\paragraph{Resubstitution error}
It measures what fraction of the class in a node is classified incorrectly if
we assign every case to the majority class in that node.
\paragraph{Gini} $i(t)=\sum_j p(j|t)(1-p(j|t))$
\paragraph{entropy}$i(t)=\sum_j p(j|t)log p(j|t)$

@@

\subsection{Missing data}
The easiest way is  to throw away records with missing data.
This has the following disadvantages: Potential bias and loss of power (which
means that our precision is reduced because we loose a lot of examples).
If we want to use a tree with incomplete data we have the following problems:
How to determine the quality of a spiet? Which way gets an observation send
with a missing value?
If the class labe misses its useless anyway so it should be thrown away.
A solution is ignoring the quality aspect and add surogate splits.
\subsection{Computational efficiency}
A categorical attribute with $L$ is distinct values has $2^{L-1}-1$ possible
distinct splits.
For binary class problems and impurity measures that belong to class $\mathcal{F}$
there is a more efficient algorithm for finding the optimal split.
$p(0|x=b_l)$  denotes the realitve frequency of class 0 for observations in the
current node with $x=b_l$. Order $p(0|x=b_l)$. Then one of the $L$ subsets
is the optimal split. Thus the search is reduced to $O(L-1)$
\subsubsection{Tree construction}
   Asuming all $p$ attributes $x_1, ..., x_p$ are numeric we have to sort
   them per attribute, which takes $O(n log n)$ time.
   Best case (where each split is a perfect subdivision) we traverse the tree
   and have to consider $O(n)$ splits per level resulting in $O(n log n)$ time.
   Worst case it becomes $O(n^2)$
\section{Graphical models}
\subsection{Probability rules}
sum rule $P(X)=\sum_y P(X,Y)$, product rule $P(X,Y)=P(X|Y)P(Y)$.
To estimate the joint distrubtion of $X$ and $Y$ we use $\overset{\wedge}{P}(x,y)=\frac{n(x,y)}{n}$
\subsection{Independence}
Random vectors $X$ and $Y$ are independent iff $P(x,y)=P(x)P(y) \text{ for all } (x,y)$
Therefore $P(x|y)=P(x)$. Independence can be written as $X \indep Y$.
This gives the factorisation criterion for independent randomvectors:
random vectors $X$ and $Y$ are independent iff there exists two functions $g$
and $h$ such that: $P(x,y) = g(x)h(y)$ for all $(x,y)$.
Random vectors $X$ and $Y$ re independent given $Z$ iff
$P(x,y,z)=P(x,z)P(y,z)/P(z)$.
\section{Baysian networks}
\section{Freq item set mining}
\end{multicols}
\end{document}

>>>>

\begin{center}
     \Large{\underline{Datamining}} \\
\end{center}
\section{Classification trees}
\subsection{Building classification trees}
The objective of tree construction is to finally obtain nodes that are pure,
in the sense that they contain cases of a single class only.
\subsubsection{Impurity and quality of a split}
Impurity is a function:
\[i(t)=\phi(p_1,p_2,...,p_J)\]
where $p_j(j=1,...,J)$ is the relative frequency of the $J$ different classes.
$\phi$ is the impurity function with the following constraints:
\begin{enumerate}
  \item An impurity measure of a node should be at maximum when the observations
    are distributed evenly over all classes in that node.
  \item It should be at a minimum when all observations belong to a single class.
  \item $\phi$ is a symmetric function of $p_1,...,p_J$.
    a symmetric function of n variables is one whose value at any n-tuple of
    arguments is the same as its value at any permutation of that n-tuple.
\end{enumerate}
The quality of a split is defined as follows:
\[\Delta i(s,t)=i(t)-\phi(l)i(l)-\phi(r)i(r)\]
Where $\phi$ is a function that determines the proportion of cases for a node
and $l$ and $r$ are the left and right nodes of $t$.
\paragraph{Resubstitution error}
It measures what fraction of the class in a node is classified incorrectly if
we assign every case to the majority class in that node.
\paragraph{Gini} $i(t)=\sum_j p(j|t)(1-p(j|t))$
\paragraph{entropy}$i(t)=\sum_j p(j|t)\log p(j|t)$

@@

\subsection{Missing data}
The easiest way is  to throw away records with missing data.
This has the following disadvantages: Potential bias and loss of power (which
means that our precision is reduced because we loose a lot of examples).
If we want to use a tree with incomplete data we have the following problems:
How to determine the quality of a spiet? Which way gets an observation send
with a missing value?
If the class labe misses its useless anyway so it should be thrown away.
A solution is ignoring the quality aspect and add surogate splits.
\subsection{Computational efficiency}
A categorical attribute with $L$ is distinct values has $2^{L-1}-1$ possible
distinct splits.
For binary class problems and impurity measures that belong to class $\mathcal{F}$
there is a more efficient algorithm for finding the optimal split.
$p(0|x=b_l)$  denotes the realitve frequency of class 0 for observations in the
current node with $x=b_l$. Order $p(0|x=b_l)$. Then one of the $L$ subsets
is the optimal split. Thus the search is reduced to $O(L-1)$
\subsubsection{Tree construction}
   Asuming all $p$ attributes $x_1, ..., x_p$ are numeric we have to sort
   them per attribute, which takes $O(n \log n)$ time.
   Best case (where each split is a perfect subdivision) we traverse the tree
   and have to consider $O(n)$ splits per level resulting in $O(n \log n)$ time.
   Worst case it becomes $O(n^2)$
\section{Graphical models}
\subsection{Probability rules}
sum rule $P(X)=\sum_y P(X,Y)$, product rule $P(X,Y)=P(X|Y)P(Y)$.
To estimate the joint distrubtion of $X$ and $Y$ we use $\overset{\wedge}{P}(x,y)=\frac{n(x,y)}{n}$
\subsection{Independence}
Random vectors $X$ and $Y$ are independent iff $P(x,y)=P(x)P(y) \text{ for all } (x,y)$
Therefore $P(x|y)=P(x)$. Independence can be written as $X \indep Y$.
This gives the factorization criterion for independent random vectors:
random vectors $X$ and $Y$ are independent iff there exists two functions $g$
and $h$ such that: $P(x,y) = g(x)h(y)$ for all $(x,y)$.
A ``log version'' is: $\log P(x,y)=g'(x)+h'(y)$ for all $(x,y)$.
Random vectors $X$ and $Y$ re independent given $Z$ iff
$P(x,y|z)=P(x|z)P(y|z)$ for all $(x,y)$ and for all $z$ for which $P(z)>0$.
This is the same as:
$P(x,y,z)=P(x,z)P(y,z)/P(z)$.
With marginal independence we can state a simple factorization criterion
to establish conditional independence: random vectors $X$ and $Y$ are
conditionally independent given $Z,X \indep Y | Z$ iff there exists
functions $g$ and $h$ such that $P(x,y,z)=g(x,z)h(y,z)$.
\subsection{Independence graphs}
Conditional independence relations can be represented as a conditional
independence graph. Let $X=(X_1,X_2,\dots, X_k)$ be a $k$-dimensional random
vector. The conditional independence graph of $X$ is the undirected graph
$G=(K,E)$ with $K={1,2,...,k}$ and where ${i,j}$ is not in the edge set $E$ 
iff $X_i \indep X_j|rest$.
Global markov property: $a$ seperates $b$ from $c$ ($a,b,c$ disjoint) iff
$X_b \indep X_c | X_a$ where $X_a=(X_i;i\in a)$ and $a$ seperates $b$ from
$c$ if for all $i \in b, j\in c:$ $a$ seperates $i$ from $j$.
Is equal to local markov property: $X_i \indep$ rest | boundary$(i)$ where
the boundary of a vertex $i$ is simply the set of adjecent vertices.
\subsection{Log linear models}
\subsubsection{for binary data}
A random experiment that only disguises between two possible outcomes is
called a \emph{Bernoulli} experiment. Random variable $X$ denotes the number of
successes in a Bernoulli experiment; $X$ therefore has possible values 0 and 1.
The probability distribution of $X$ is completely determined by the probablity of
success which we denote by $p$  and is $P(X=0) = 1 - p $ and $ P(X=1)=p$.
$X$ has the probability function $P(x)=p^x(1-p)^{1-x}$ for $x=0,1$ and
$0 \leq p \leq 1$.
log properities: $\log a^b=b \log a$, $\log ab = \log a + \log b$,
$\log \frac{a}{b}=\log a - \log b$.
The cartesion product 
${0,1} \times {0,1}$ creates a table of the bivariate Bernoulli random
vector $(X_1,X_2)$.
\begin{tabular}{llll}
  $P(x_1,x_2)$ & $x_2=0$ & $x_1=1$ & Total \\
  $x_1=0$ & $p(0,0)$ & $p(0,1)$ & $p_1(0)$ \\
  $x_1=1$ & $p(1,0)$ & $p(1,1)$ & $p_1(1)$ \\
  Total & $p_2(0)$ & $p_2(1)$ & $1$ \\
\end{tabular}
which can be written as
$P(x_1,x_2)=p(0,0)^{(1-x_1)(1-x_2)}p(0,1)^{(1-x_1)x_2}p(1,0)^{x_1(1-x_2)}p(1,1)^{x_1x_2}$
for $x_1=0,1$ and $x_2=0,1$.
Taking logarithms of this identity for $P$ and collecting terms in$x_1$ and
$x_2$ gives:
$\log P(x_1,x_2)= \begin{matrix}
    \log p(0,0) + x_1 \log \frac{p(1,0)}{p(0,0)} +  \\
    x_2 \log \frac{p(0,1)}{p(0,0)} + x_1 x_2 \log \frac{p(1,1)p(0,0)}{p(0,1)p(1,0)}
    \end{matrix}
$
reparameterizing the right hand side leads tothe so-caled \emph{log-linear expension}
$\log P(x_1,x_2)=u_\emptyset+ x_1u_1 + x_2 u_2 + x_1 x_2 u_{12}$
the $u$ coefficients are called $u$-terms, they substitue the log part.
The coefficient of the product $x_1x_2$ is the logarithm of the cross product
ratio:
$u_{12}=\log cpr(X_1,X_2)$
The random variables $X_1$ and $X_2$ are independent iff $u_{12}=0$
If $u_{12}=0$ we can take $g(x_1)=u_\emptyset + x_1 u_1$ and $h(x_2)= u_2 x_2$.
This is not posible with $u_{12}\neq 0$.
We can enforce (conditional) independence constraints by etting the right
$u$-terms to zero.
\subsection{non binary data}
To work with non binary data we make the u terms functions rather than constants:
$log P(x_1,x_2)=u_\emptyset + u_1(x_1) + u_2(x_2) + u_{12}(x_1,x_2)$
This produces to many parameters, and to identify them we have to impose
an extra constraint: $u_a(x_a)=0$ whenever $x_i=0$ and $i\in a$. Here we
assume that if $x_i$ has $d_i$ possible values, these are numbered
$0,1,...,d_i-1$, but not ordered.
\subsection{Hierarchical and graphical}
% TODO: perhaps delete this section?
The log-linear expension of the crossclassified multinominal density function
$P_K$ is $\log P_K(x)=\underset{a \subseteq K}{\sum} u_a(x_a)$ where the sum is
taken over all possible subsets of $K={1,2,...,k}$ and where the $u$-terms are
coorinate projection functions, so that $u_a(x)=u_a(x_a)$ and also satisfy the
constraint that $u_a(x)=0$ whenever $x_i = 0$ and $i\in a$.
\subsubsection{Independence and the u-terms}
$If(X_a,X_b,X_c)$ is a partitioned multinominal random vector then
$X_b \indep X_c|X_a$ if and only if all u-terms in the log-linear expension with
one or more coordinate in $b$ and one or more coordinate in $c$ are zero.
The proof is direct application of the factorisation theorom for conditional
independence.
\subsubsection{Graphical model}
Given an independence graph $G=(K,E)$ the cross-classified multinomal
distrubtion for the random vector $X$ is a graphical model for $X$ if the
distribution of $X$ is an arbitrary apart from constraints of the form that for
all pairs of coordinates not in the edge set $E$ of $G$ the $u$ terms containing
the selected coordinates are identically zero.
\subsection{Maximum likelhood estimation}
The maximum likelihood estimator of gaphical log linear model $M$ satisfies the
likelihood equations: $\overset{\wedge}{n}_a=N\overset{\wedge}{P}_a=n_a$
\section{Baysian networks}
\section{Freq item set mining}
\end{multicols}
\end{document}
